\ProjectName{}: Data Imputation as a Query Optimization

### 2016-12-12 paper/algorithm.tex
paper: Add to examples section
paper: edit loss section and rearrange related work

The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in any modern RDBMS.\@ Our runtime cost model is perhaps less sophisticated than most, as it is identical to the cost model in SimpleDB.

\todo{Check loss calculation w/ Micah}
$\text{Loss}(Q)$ is computed as follows.
For each sub-query, in addition to a histogram, we keep track of the estimated number of null values in each column.
\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{Null}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{Null}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss penalty of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $0 < p \leq 1$ which decreases as the number of tuples available to train the imputation algorithm increases.
The intuition behind this formula is that imputation performance should increase if there is more training data available, but there is a saturation point beyond which more data provides little additional benefit.

@@

The dirty set over-approximates the set of attributes that contain null. For example, a filter might remove all tuples which contain null without changing the dirty set, forcing an unnecessary imputation. We choose to over-approximate the dirty set to avoid the possibility of dropping a tuple that contains a null value without explicitly imputing the value or applying a drop operator.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data} and others) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the researcher to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}.

@@

>>>>

The computation of $\text{Time}(Q)$ is standard and is comparable to the cost model used in
any modern RDBMS.\@ Our runtime cost model is the cost model used in SimpleDB, though modern systems
have more sophisticated approaches.

$\text{Loss}(Q)$ is computed as follows.
For each sub-query, in addition to a histogram, we keep track of the estimated number of
null values in each column, $\text{MissingCount}(c, Q)$.
\begin{align*}
  \text{Loss}(Q) = \begin{cases}
    \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \delta_C(Q') \\
    \frac{1}{\sqrt{|Q'|}} \sum_{c \in C} \text{MissingCount}(c, Q') & Q = \mu_C(Q') \\
    \text{Loss}(Q_1') + \text{Loss}(Q_2') & Q = Q_1' \bowtie_\psi Q_2' \\
    \text{Loss}(Q') & Q = \sigma_\phi(Q'), Q = \pi_C(Q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a first approximation to the learning guarantees of
regression trees. \todo{Micah: Probably need to look into this more carefully since I don't
know much about the learning guarantees of regression trees!} Since we conceive that
alternate imputation strategies could be used, the analyst would provide a corresponding
loss with a reasonable parameterization.

\todo{Micah: Notation not super clear to me. I'm guessing $|Q'|$ is the number of tuples total, but
$Q'$ is a query plan, and I would interpret it's cardinality as perhaps the height of the
tree or number of elements in the tree, not in terms of underlying relation. Similarly, I'm
not sure how to define $\text{MissingCount}(\cdot)$ above.}

@@

That is, the dirty set over-approximates the set of attributes that contain null. For example, a filter might remove all tuples which contain null without changing the dirty set, forcing an unnecessary imputation. We choose to over-approximate the dirty set to avoid the possibility of dropping a tuple that contains a null value without explicitly imputing the value or applying a drop operator.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data} and others) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}.

@@

The flexibility of this non-parametric imputation approach aligns well with the spirit of
query optimization, which aims to reconcile performance with a declarative interface to data
manipulation. By removing concerns for imputation, users executing queries in \ProjectName{}
don't need to consider the potential nature of the missing data.
